---
title: "Bats_DEseq_without3"
output: html_document
---
```{r}
#import library to be able to build table count for each gene
library("tximport")
library(DESeq2)
library(biomaRt) #to automatically request ensembl identifiers
library(dplyr) #for the manipulation of data frames
library(tidyr)
library(data.table)
#set working directory
setwd("/ifb/data/mydatalocal/data_from_ftp")
```

Import the files, we need a variable that contains the name of the files quant.sf
```{r}
dir <- "/ifb/data/mydatalocal/data_from_ftp/outputs_salmon" #the directory where we look for the files
all_files <- list.files(dir)

#get the quant.sf files for all samples
#initialisation
files_Lib <- c()
names_file <- c()
i <- 1

#loop to get all the files in the same place
for (name in all_files) {
    if (grepl("salmon_Lib",as.character(name))==TRUE) {
      files_Lib[i] <-  file.path(dir,name,"quant.sf", fsep="/")
      names_file[i] <- c(name)
      i <- i+1
      }
  }

names(files_Lib) <- names_file #give them names

file_wo3 <- files_Lib[-3] #Remove Lib3 (CTL that was an outlier)
```

Now we build reading table from the trinity gene trans map, to have the correspondance between gene and transcript
```{r}
Tab_trini <- read.table("/ifb/data/mydatalocal/data_from_ftp/trinity_data/Trinity_RF.fasta.gene_trans_map") #get the fasta.gene_trans_map
names(Tab_trini) <- c("GENEID","TXNAME") #to have the right column names
Tab_trini <- Tab_trini[,c(2,1)] #to have the right order: TXNAME as first column, GENEID as second
```

Now let's build a table with the count from each gene from the outputs of salmon 
```{r}
txi_wo3 <- tximport(file_wo3, type = "salmon", tx2gene = Tab_trini)
names(txi_wo3) #outputs, we obtained 3 tables: abundance, counts, and length
head(txi_wo3$counts) #what's interesting for DESeq is the counts
```

Now let's use DESeq
```{r}
#build table of correspondence wich sample is + or - IFN
samp.name_wo3 <- colnames(txi_wo3$counts)
samp.type_wo3 <- factor(c("CTL","CTL","IFN","IFN","IFN")) #2 first are control, 3 last are +IFN
samples_wo3 <- data.frame(run=samp.name_wo3,condition=samp.type_wo3)

#now build DESeq object (dds)
ddsTxi_wo3 <- DESeqDataSetFromTximport(txi_wo3, colData = samples_wo3, design = ~ condition)

#Run DEseq
dds_wo3 <- DESeq(ddsTxi_wo3)
res_wo3 <- results(dds_wo3)
#res_wo3
#save the table
#write.table(res_wo3,"/ifb/data/mydatalocal/data_from_ftp/DESeq/DESeqRes_withoutLib3.tab")

#count nb of significative p-values
table(res_wo3$padj <0.05) #nb of tests that are under the threshold of 0.05 #1670
```
Now we do control quality, and verifiy that there is no abnormality with the replicates (samples from the same conditions have to be closer than different conditions)

```{r}
#MA-plot 
plotMA(res_wo3, ylim=c(-2,2))

#PCA
vsd_wo3 <- vst(dds_wo3, blind=FALSE)
plotPCA(vsd_wo3, intgroup=c("condition"))
plotPCA(vsd_wo3, intgroup=c("condition","run"))
#Now the conditions are grouped together.
```

Now we want to cross the results of Blast with the outputs from DEseq, in order to identify the names of all the genes (based on their corresponding human CDS)
```{r}
#Import Blast table
Blast <- read.table("/ifb/data/mydatalocal/data_from_ftp/blast_data/DataBats.blast")

#now we make the correspondence between ENSEMBL identifiers and external human gene names, using biomaRt
#match gene/transcript names
ensembl <- useEnsembl(biomart = "genes", dataset = "hsapiens_gene_ensembl")
tx2geneHomo = biomaRt::getBM(attributes=c('ensembl_transcript_id', 
                                      'ensembl_gene_id', 'external_gene_name'),mart=ensembl)
#add gene name to the blast outputs
blastHomo<-data.table(Blast)
blastHomo<-blastHomo[, c("trid", "version") := tstrsplit(V2, ".", fixed=TRUE)][]
blastHomo<-blastHomo[, c("gene", "alttr") := tstrsplit(V1, "_i", fixed=TRUE)][]

blastHomo1 <- blastHomo%>% group_by(gene)%>% slice_max(order_by = V12, n = 1, with_ties = TRUE)

blastHomoName<-merge(tx2geneHomo,blastHomo1,by.x="ensembl_transcript_id",by.y="trid") #to merge the 2 dataframe, one with the Blast output and ensembl names (blastHomo1), and one with the ensemble and external gene names (tx2gneHomo)

blastHomoNameUniq<-unique(blastHomoName[,c("ensembl_gene_id","external_gene_name","gene")]) #unique, to remove the duplicates

#associate our DE results with gene names from human
resUsName_wo3=merge(as.data.frame(res_wo3),blastHomoNameUniq,by.x=0,by.y="gene")
#view(ResUsName_wo3)
#save this table
write.csv(resUsName_wo3,file="/ifb/data/mydatalocal/data_from_ftp/DESeq/Tab_foldchange_without3.csv")
```

Number of genes up and down regulated
```{r}
table(resUsName_wo3$padj <0.05) #Number of genes identified as significatively differently expressed (with interferon vs without) #696
table(resUsName_wo3$padj <0.05 & resUsName$log2FoldChange>0) #Number of genes significatively UPregulated #521
table(resUsName_wo3$padj <0.05 & resUsName$log2FoldChange<0) #Number of genes significatively DOWNregulated #175

#filter the downregulated and upregulated genes
UpReg <- filter(resUsName_wo3, padj<0.05 & resUsName$log2FoldChange>0)
DownReg <- filter(resUsName_wo3, padj<0.05 & resUsName$log2FoldChange<0)
AllReg <- filter(resUsName_wo3, padj<0.05)
```

```{r}
#Genes in common with or withoit Lib3
#import our table without 
resWithLib3 <-read.csv("/ifb/data/mydatalocal/data_from_ftp/DESeq/Tab_foldchange_genes.csv")

CommonGenesW3 <- merge(resUsName_wo3,resWithLib3,by.x="external_gene_name", by.y="external_gene_name")

#Chi2 test

cont_tb_wo3<-table(withoutlib3=CommonGenesW3$padj.x<0.05,withlib3=CommonGenesW3$padj.y < 0.05) #Contingency table
chisq.test(cont_tb_wo3) #

#add a column for our test to mark if gene is up-regulated, down-regulated or not regulated
CommonGenesW3$WithLib3="nonreg" #by default
CommonGenesW3$WithLib3[CommonGenesW3$padj.x<0.05&CommonGenesW3$log2FoldChange.x>0]="Up"
CommonGenesW3$WithLib3[CommonGenesW3$padj.x<0.05&CommonGenesW3$log2FoldChange.x<0]="Down"
#idem for Holzer's test
CommonGenesW3$WithoutLib3="nonreg"
CommonGenesW3$WithoutLib3[CommonGenesW3$padj.y<0.05&CommonGenesW3$log2FoldChange.y>0]="Up"
CommonGenesW3$WithoutLib3[CommonGenesW3$padj.y<0.05&CommonGenesW3$log2FoldChange.y<0]="Down"

cont_tb2_wo3 <- table(WithLib3=CommonGenesW3$WithLib3,WithoutLib3=CommonGenesW3$WithoutLib3) #Contingency table 2, for all groups
chisq.test(cont_tb2_wo3) #
```



